---
title: "IST707_FinalProject"
output: pdf_document
---
```{r}
#importing in necessary libraries:
library(dplyr)
library(grid)
#library(tm)
library(arules)
library(arulesViz)
library(tidyverse)
library(e1071)
library(caret)
```

```{r}
#loading the data set:
heart <-read_csv("heart.csv")
summary(heart)
```
```{r}
#examinging structure:
str(heart)
```
```{r}
#checking for missing values:
sum(is.na(heart))
```
```{r}
#checking for duplicates:
sum(duplicated(heart))
```
There is 1 duplicate record
```{r}
#removing duplicates:
heart <-heart[!duplicated(heart),]
```
```{r}
str(heart)
```
Removing missing values and duplicates, the data set is now 302 records with 14 observations

Converting Variables from numeric to nominal
```{r}
heart$sex <-factor(heart$sex)
heart$cp <-factor(heart$cp)
heart$fbs <-factor(heart$fbs)
heart$restecg <-factor(heart$restecg)
heart$exang <-factor(heart$exang)
heart$slope <-factor(heart$slope)
heart$ca <-factor(heart$ca)
heart$thal <-factor(heart$thal)
heart$target <-factor(heart$target)
```
```{r}
#checking:
str(heart)
```

Recoding variables and discretizing variables: 
```{r}
#age:
heart$age <-cut(heart$age, breaks= c(0,20,30,40,50,60,70,80,90),
               labels=c("teens","twenties","thirties","forties","fifties","sixties","seventies","eighties"))
#sex:
heart$sex <-recode(heart$sex,"0"="Female","1"="Male")
#cp:
heart$cp <-recode(heart$cp, "0"="typical_angina","1"="atypical_angina","2"="non-anginal_pain","3"="asymptomatic")
#trestbps:
heart$trestbps <-cut(heart$trestbps, breaks = c(0,120,140,160,180,200),
                     labels=c("optimal","prehypertension","high blood pressure stage 1","high blood pressure stage 2","hypertension crisis"))
#chol:
heart$chol <-cut(heart$chol, breaks= c(0,200,240,600),
                 labels=c("Healthy Chol","Borderline","High Risk"))
#fbs:
heart$fbs <-recode(heart$fbs, "0"="Less than 120mg/dl fasting blood sugar", "1"="More than 120mg/dl fasting blood sugar")

#restecg:
heart$restecg <-recode(heart$restecg,"0"="Normal","1"="ST-T_wave_abnormality","2"="Probable/Definite_left_ventricular_hypertrophy")

#thalach:
heart$thalach <-cut(heart$thalach, breaks= c(0,100,150,250),
                 labels=c("low","normal","high"))


#exang:
heart$exang <-recode(heart$exang, "0"="No","1"="Yes")

#oldpeak:
heart$oldpeak <-ifelse(heart$oldpeak==0,"Zero","More than Zero")

#slope:
heart$slope <-recode(heart$slope,"0"="upsloping","1"="flat","2"="downsloping")

#ca:
heart$ca <-factor(heart$ca)

#thal
heart$thal <-recode(heart$thal,"0"="absence","1"="normal","2"="fixed_defect","3"="reversable_defect")

#target:
heart$target <-recode(heart$target,"0"="low risk", "1"="high risk")

#checking:
str(heart)
```

Exploratory Visualizations:
```{r}
plot_age<-heart%>%
  count(age)%>%
  ggplot(aes(age, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=5)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("age")
plot_age

plot_sex<-heart%>%
  count(sex)%>%
  ggplot(aes(sex, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("sex")
plot_sex

plot_cp<-heart%>%
  count(cp)%>%
  ggplot(aes(cp, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=3)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("cp")
plot_cp

plot_trestbps<-heart%>%
  count(trestbps)%>%
  ggplot(aes(trestbps, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=4)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("trestbps")
plot_trestbps

plot_chol<-heart%>%
  count(chol)%>%
  ggplot(aes(chol, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=3)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("chol")
plot_chol

plot_fbs<-heart%>%
  count(fbs)%>%
  ggplot(aes(fbs, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("fbs")
plot_fbs

plot_restecg<-heart%>%
  count(restecg)%>%
  ggplot(aes(restecg, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("restecb")
plot_restecg
```

```{r}
plot_thalach<-heart%>%
  count(thalach)%>%
  ggplot(aes(thalach, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white")+
  #geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("thalach")
plot_thalach

plot_exang<-heart%>%
  count(exang)%>%
  ggplot(aes(exang, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("exang")
plot_exang

plot_oldpeak<-heart%>%
  count(oldpeak)%>%
  ggplot(aes(oldpeak, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("oldpeak")
plot_oldpeak

plot_slope<-heart%>%
  count(slope)%>%
  ggplot(aes(slope, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=3)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("slope")
plot_slope

plot_ca<-heart%>%
  count(ca)%>%
  ggplot(aes(ca, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("ca")
plot_ca

plot_thal<-heart%>%
  count(thal)%>%
  ggplot(aes(thal, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=4)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("thal")
plot_thal

plot_target<-heart%>%
  count(target)%>%
  ggplot(aes(target, y=n))+
  geom_bar(stat="identity",fill="cornflowerblue",col="white",bins=2)+
  geom_text(aes(label=n),position=position_dodge(width=0.9),vjust=-0.25)+
  ylab("count")+ xlab("target")
plot_target


```
Correlation;
```{r}
library(corrplot)
heart2 <-read_csv("heart.csv")
cor_heart <-cor(heart2)
round(cor_heart,2)
par(mfrow=c(1,1))
corrplot(cor_heart, method='number',tl.srt=45,number.cex=0.5,type='upper',order='hclust')
heart2[,10]
```

plots with just significant attributes:
```{r}
#positively correlated
cp_target <-table(heart$cp,heart$target)
barplot(cp_target, beside=TRUE,col=c("cornflowerblue","orchid2","coral2","seagreen"),
        legend.text=rownames(cp_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs CP")

thalach_target <-table(heart$thalach,heart$target)
barplot(thalach_target, beside=TRUE,col=c("cornflowerblue","orchid2","coral2"),
        legend.text=rownames(thalach_target),args.legend=list(x="topright",cex=0.5),
         main="Heart Attack Risk vs Thalach")

slope_target <-table(heart$slope,heart$target)
barplot(slope_target, beside=TRUE,col=c("cornflowerblue","orchid2","coral2"),
        legend.text=rownames(slope_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs Slope")

#negatively correlated
exang_target <-table(heart$exang,heart$target)
barplot(exang_target, beside=TRUE,col=c("cornflowerblue","orchid2"),
        legend.text=rownames(exang_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs Exang")

oldpeak_target <-table(heart$oldpeak,heart$target)
barplot(oldpeak_target, beside=TRUE,col=c("cornflowerblue","orchid2"),
        legend.text=rownames(oldpeak_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs Old Peak")

ca_target <-table(heart$ca,heart$target)
barplot(ca_target, beside=TRUE,col=c("cornflowerblue","orchid2","coral2","seagreen","dodgerblue"),
        legend.text=rownames(ca_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs CA")

sex_target <-table(heart$sex,heart$target)
barplot(sex_target, beside=TRUE,col=c("cornflowerblue","orchid2"),
        legend.text=rownames(sex_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs Sex")

thal_target <-table(heart$thal,heart$target)
barplot(thal_target, beside=TRUE,col=c("cornflowerblue","orchid2","coral2","seagreen"),
        legend.text=rownames(thal_target),args.legend=list(x="topright",cex=0.5),
        main="Heart Attack Risk vs Thal")
```

Modeling:

## Association Rule Mining (ARM)
```{r}
str(heart)
heart[,8]
#oldpeak is char. Converting to factor.
heart$oldpeak <- as.factor(heart$oldpeak)

#Confirming
str(heart$oldpeak)

heartID <- as.character(heart[["id"]])
heart$heartID <- NULL
heartTransactions <- as(heart, "transactions")
str(heartTransactions)
#View item frequency plot
itemFrequencyPlot(heartTransactions, topN=20, type="absolute")
```

**Item Frequency Plot for High Risk Patients**
```{r}
heartHighRisk <- heart %>%
  filter(target=="high risk")

HighRiskTransactions <- as(heartHighRisk, "transactions")
#View item frequency plot
itemFrequencyPlot(HighRiskTransactions, topN=20, type="absolute")
```


### Apriori Algorithm

**Initial Attempt:**
```{r}
#supp = .005, conf = .8, minlen = 3
heartRulesInitial <- apriori(heartTransactions, parameter=list(supp=.005, conf=.8, minlen=3),
                      appearance=list(default="lhs", rhs="target=high risk"),
                 control=list(verbose=F))

heartRulesInitialByLift <- sort(heartRulesInitial, decreasing=TRUE, by="lift")
options(digits=2)
heartRulesInitialByLift
inspect(heartRulesInitialByLift[1:10])

```
```{r}
#Summary of initial attempt
summary(heartRulesInitialByLift)
```



**Attempt 2:**
```{r}
#supp = .02, conf = .85, minlen = 3
heartRules2 <- apriori(heartTransactions, parameter=list(supp=.02, conf=.85, minlen=3),
                      appearance=list(default="lhs", rhs="target=high risk"),
                 control=list(verbose=F))

heartRules2LiftSort <- sort(heartRules2, decreasing=TRUE, by="lift")
options(digits=2)
heartRules2LiftSort
inspect(heartRules2LiftSort[1:10])
```
```{r}
#Summary of attempt 2
summary(heartRules2LiftSort)
```



**Final Attempt:**
```{r}
heartRules3 <- apriori(heartTransactions, parameter=list(supp=.11, conf=.95, minlen=3),
                      appearance=list(default="lhs", rhs="target=high risk"),
                 control=list(verbose=F))

summary(heartRules3)
```

**Inspect All Strong Rules**
```{r}
inspect(heartRules3)
```

**Strong Rules Sorted By Support**
```{r}
heartRules3SupSort <- sort(heartRules3, decreasing=TRUE, by="sup")
options(digits=2)
heartRules3SupSort
inspect(heartRules3SupSort[1:10])
```

**Strong Rules Sorted By Confidence**
```{r}
heartRules3ConfSort <- sort(heartRules3, decreasing=TRUE, by="conf")
options(digits=2)
heartRules3ConfSort
inspect(heartRules3ConfSort[1:10])
```

**Strong Rules Sorted By Lift**
```{r}
heartRules3LiftSort <- sort(heartRules3, decreasing=TRUE, by="lift")
options(digits=2)
heartRules3LiftSort
inspect(heartRules3LiftSort[1:10])
```

**Plot Rules**
```{r}
plot(heartRules3LiftSort, method="graph")
```

# Classification Models
# Naive Bayes
```{r}
trainPartition <- createDataPartition(heart$target, p=.7, list=FALSE)

trainHeart <- heart[trainPartition, ]
testHeart <- heart[-trainPartition, ]
testHeart <- na.omit(testHeart)

dim(trainHeart)
dim(testHeart)

table(trainHeart$target)
table(testHeart$target)
```

## Initial Naive Bayes Attempt
```{r}
set.seed(125)
nbTrain <- naiveBayes(target~., data=trainHeart)

#See training model results including conditional probabilities
nbTrain
```


```{r}
nbPredict <- predict(nbTrain, testHeart, type="class")
confusionMatrix(nbPredict, testHeart$target)
```
```{r}
cbind(nbPred, testHeart)
nbPred <- as.data.frame(nbPredict)
```


```{r}
plot(nbPredict, ylab = "Density", main = "NaiveBayes Plot", col="darkred", ylim = c(0,60))
actual<- table(testHeart$target)
barplot(actual, col = "darkred", ylab = "Density", main = "Test Dataset Plot", ylim = c(0,60))


```



